Approximating the Number of COVID-19 Infections in the United States Across 2021

### 2023-07-18 writing/manuscript.Rmd
update manuscript & supplement

As a result of these dramatic changes, Chitwood *et al.*@chitwood2022 made substantial changes to Covidestim, as noted in the model change log. Because the variant causes much milder infections, the infection fatality for Omicron infections is lower than previous variants, and death counts were much lower relative to the number of infections. To handle this change, rather than fitting model with deaths, they switched to using hospitalizations. They also allowed for the possibility of reinfections, since although reinfections were more rare with previous variants, Omicron is associated with higher reinfection rates [@pulliam2022]. The changes in the model may contribute to the differences we see between the probabilistic bias intervals and Covidestim intervals during the Omicron wave.

@@

In Figure (CHOOSE BETWEEN VERSION BY COLOR AND FACETED), we present simulation intervals for each implementation of probabilistic bias analysis.

In the first three implementations, the prior distributions specified for the bias parameters vary by state and date. In particular, we use survey data from the COVID-19 Trends and Impact Survey [@salomon2021a] to inform the prior distributions for two of the bias parameters: the probability of having symptoms among the untested population ($\Pr(S_1|\text{untested})$), and the parameter $\beta$, which represents the ratio of the test positivity among members of the untested population who have no symptoms to the overall test positivity. More detail on these implementations is provided in the [Methods](#Methods) section. In the last implementation, the prior distributions are informed by the distributions from the survey data across all dates and states, but the same prior distributions are used for all states and time intervals considered. While there is substantial overlap among all of the versions, we see the implementation allowing $\beta$ and $\Pr(S_1|\text{untested})$ to vary by state and date tends to be the highest, followed by that only allowing $\beta$ to vary by state and date,  followed by that only allowing $\Pr(S_1|\text{untested})$ to vary by state and date, and lastly the implementation that does not vary the priors by state or date tends to produce the lowest estimates.

```{r sim-intervals-faceted-county, fig.width=15, fig.height=16,  fig.cap = "Simulation intervals for counties in Massachusetts. Each interval corresponds to a 95\\% simulation interval for the total number of estimated infections for that county in that two-week time interval. The columns represent different implementations of the probabilistic bias analysis: in the first through the third columns, the priors vary by state and date, while for the fourth, they are the same for all states and time intervals considered. For the implementation in the first column, we center the distribution of $\\beta$ at the ratio of the screening test positivity to the overall test positivity from the survey, and we center the distribution of $\\Pr(S_1|\\text{untested})$ at the percentage of the population experiencing COVID-19-like illness from the survey for each two-week interval. The second column centers only $\\beta$ at the aformentioned value, and the third column only centers $\\Pr(S_1|\\text{untested})$ at the aforementioned value. The fourth column corresponds to the implementation where we specify priors that are the same for all dates. The implementation that centers both $\\Pr(S_1|\\text{untested})$ and $\\beta$ at the survey values is consistently the highest among the implementations, followed by the implementation that centers only $\\beta$ at the survey value, followed then by the implementation only centering $\\Pr(S_1|\\text{untested})$ at the survey value, and then the lowest among the implementations is that where the priors do not vary by date."}

@@

```{r sim-intervals-color-county, fig.width=15,  fig.cap = "Simulation intervals for counties in Massachusetts, colored by the implementation of probabilistic bias analysis. In the first three implementations, the priors vary by state and date, while for the fourth, they are the same for all states and time intervals considered. For the first implementation, we center the distribution of $\\beta$ at the ratio of the screening test positivity to the overall test positivity from the survey, and we center the distribution of $\\Pr(S_1|\\text{untested})$ at the percentage of the population experiencing COVID-19-like illness from the survey for each two-week interval. The second implementation, we center only $\\beta$ at the aformentioned value, and the for the next we only center $\\Pr(S_1|\\text{untested})$ at the aforementioned value. The last implementation corresponds to the implementation where we specify priors that are the same for all dates. The implementation that centers both $\\Pr(S_1|\\text{untested})$ and $\\beta$ at the survey values is consistently the highest among the implementations, followed by the implementation that centers only $\\beta$ at the survey value, followed then by the implementation only centering $\\Pr(S_1|\\text{untested})$ at the survey value, and then the lowest among the implementations is that where the priors do not vary by date."}

include_graphics(here('figures/simulation-intervals-color-version-county.pdf'))

@@

Also of interest beyond the total numbers of infections is the ratio of the infections that are estimated to those that were observed, as this gives us a sense for the case ascertainment at different time periods in the pandemic. We see in Figure \@ref(fig:ma-heatmap) that the two-week interval estimated to have the highest ratio of estimated to observed infections was July 2, 2021 through July 30, 2021, during the delta wave. This corresponds to a similar time frame that we saw at the state-level in Figure \@ref(fig:ratio-estimated-to-observed-by-wave). Hamsphire County and Suffolk County were consistently among the lowest with regard to the ratio of estimated to observed infections, which may be explained by the screening testing occurring at universities in these counties.

>>>>

As a result of these dramatic changes, Chitwood *et al.* @chitwood2022 made substantial changes to Covidestim, as noted in the model change log. Because the variant causes much milder infections, the infection fatality for Omicron infections is lower than previous variants, and death counts were much lower relative to the number of infections. To handle this change, rather than fitting model with deaths, they switched to using hospitalizations. They also allowed for the possibility of reinfections, since although reinfections were more rare with previous variants, Omicron is associated with higher reinfection rates [@pulliam2022]. The changes in the model may contribute to the differences we see between the probabilistic bias intervals and Covidestim intervals during the Omicron wave.

@@

In Figure \@ref(fig:sim-intervals-color-county), we present simulation intervals for two of the four implementations of probabilistic bias analysis. For most time intervals, the other two implementations fall between those shown in the figure.

In the first three implementations, the prior distributions specified for the bias parameters vary by state and date. In particular, we use survey data from the COVID-19 Trends and Impact Survey [@salomon2021a] to inform the prior distributions for one to two of the bias parameters: the probability of having symptoms among the untested population ($\Pr(S_1|\text{untested})$), and the parameter $\beta$, which represents the ratio of the test positivity among members of the untested population who have no symptoms to the overall test positivity. More detail on these implementations is provided in the [Methods](#Methods) section. In another, more simple, implementation, which is the only implementation presented at the state-level, the prior distributions are informed by the distributions from the survey data across all dates and states, but the same prior distributions are used for all states and time intervals considered.

While there is substantial overlap among all of the versions, the implementation allowing $\beta$ and $\Pr(S_1|\text{untested})$ to vary by state and date tends to be the highest, followed by that only allowing $\beta$ to vary by state and date,  followed by that only allowing $\Pr(S_1|\text{untested})$ to vary by state and date, and lastly the implementation that does not vary the priors by state or date tends to produce the lowest estimates (Supplementary Figure sim-intervals-faceted-county).

```{r sim-intervals-faceted-county, fig.width=15, fig.height=16,  fig.cap = "Simulation intervals for counties in Massachusetts. Each interval corresponds to a 95\\% simulation interval for the total number of estimated infections for that county in that two-week time interval. The columns represent different implementations of the probabilistic bias analysis: in the first through the third columns, the priors vary by state and date, while for the fourth, they are the same for all states and time intervals considered. For the implementation in the first column, we center the distribution of $\\beta$ at the ratio of the screening test positivity to the overall test positivity from the survey, and we center the distribution of $\\Pr(S_1|\\text{untested})$ at the percentage of the population experiencing COVID-19-like illness from the survey for each two-week interval. The second column centers only $\\beta$ at the aformentioned value, and the third column only centers $\\Pr(S_1|\\text{untested})$ at the aforementioned value. The fourth column corresponds to the implementation where we specify priors that are the same for all dates. The implementation that centers both $\\Pr(S_1|\\text{untested})$ and $\\beta$ at the survey values is consistently the highest among the implementations, followed by the implementation that centers only $\\beta$ at the survey value, followed then by the implementation only centering $\\Pr(S_1|\\text{untested})$ at the survey value, and then the lowest among the implementations is that where the priors do not vary by date.", include=FALSE}

@@

```{r sim-intervals-color-county, fig.width=15,  fig.cap = "Simulation intervals for counties in Massachusetts, colored by the implementation of probabilistic bias analysis. Only the two implementations that were consistently the highest and lowest among the implementations are included for clarity. In the   In the first three implementations, the priors vary by state and date, while for the fourth, they are the same for all states and time intervals considered. For the first implementation, we center the distribution of $\\beta$ at the ratio of the screening test positivity to the overall test positivity from the survey, and we center the distribution of $\\Pr(S_1|\\text{untested})$ at the percentage of the population experiencing COVID-19-like illness from the survey for each two-week interval. The second implementation, we center only $\\beta$ at the aformentioned value, and the for the next we only center $\\Pr(S_1|\\text{untested})$ at the aforementioned value. The last implementation corresponds to the implementation where we specify priors that are the same for all dates. The implementation that centers both $\\Pr(S_1|\\text{untested})$ and $\\beta$ at the survey values is consistently the highest among the implementations, followed by the implementation that centers only $\\beta$ at the survey value, followed then by the implementation only centering $\\Pr(S_1|\\text{untested})$ at the survey value, and then the lowest among the implementations is that where the priors do not vary by date."}

include_graphics(here('figures/simulation-intervals-color-version-only-2-county.pdf'))

@@

Also of interest beyond the total numbers of infections is the ratio of the infections that are estimated to those that were observed, as this gives us a sense for the case ascertainment at different time periods in the pandemic. We see in Figure \@ref(fig:ma-heatmap) that the two-week interval estimated to have the highest ratio of estimated to observed infections was July 2, 2021 through July 30, 2021, during the delta wave. This corresponds to a similar time frame that we saw at the state-level in Figure \@ref(fig:ratio-estimated-to-observed-by-wave). Hampsphire County and Suffolk County were consistently among the lowest with regard to the ratio of estimated to observed infections, which may be explained by the screening testing occurring at universities in these counties.
